The Weakly Nonlinear Magnetorotational Instability in a Taylor Couette Flow

### 2016-07-13 paper/widegap_mri_paper.tex
added widegap eigenfunction and nonlinear vector figures
added widegap (h)mri matrices to paper

\title{The Weakly Nonlinear Magnetorotational Instability in a Wide-Gap Channel Flow}
\author{Clark, S.E.\altaffilmark{1}}
\author{Oishi, J.S. \altaffilmark{2,}\altaffilmark{3}}
%\author{Mac Low, M.-M.\altaffilmark{1,}\altaffilmark{2}}
\altaffiltext{1}{Department of Astronomy, Columbia University, New York, NY} 
\altaffiltext{2}{Department of Astrophysics, American Museum of Natural History, New York, NY}
\altaffiltext{3}{Department of Physics, SUNY Farmingdale}

@@

@@

Many investigations of the MRI use the ``thin gap" approximation, in which the radial extent of the fluid channel is taken to be much smaller than the radius of curvature. That is, for a channel center $r_0$ bounded by inner and outer radii $r_1$ and $r_2$, respectively, the thin gap approximation applies when $r_0 \gg (r_2 - r_1)$. The thin gap approximation simplifies the MRI equations by excluding curvature terms, because the flow through a thin gap can be taken to be approximately linear in $\phi$, i.e. Cartesian. In this work we undertake the first (to our knowledge) weakly nonlinear analysis of the MRI in the wide gap regime, where the channel width may be comparable to or larger than its distance from the center of rotation.

@@

Thus through judicious choice of cylinder rotation rates, we can set $q(r_0) = 3/2$, for quasi-Keplerian flow. Note that in the thin gap approximation we have the freedom to choose any rotation profile, as the narrow gap width imposes a linear shear (constant $q$). Our base velocity is $\mathbf{u_0} = r\Omega(r) \phihat$. We initialize an axial magnetic field $\mathbf{B_0} = B_0 \zhat$.

@@

where J is the Jacobian $J(x, y) \equiv \partial_z x \partial_r y - \partial_r x \partial_z y$. Note that in the above, $\nabla^2 x \equiv \partial_r^2 x + \partial_z^2 x + \frac{1}{r} \partial_r x$. Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are nondimensionalized, where lengths have been scaled by $r_0$, velocities by $r_0 \Omega_0$, densities by $\rho_0$, and magnetic fields by $B_0$; where $B_0$ appears in the above it is formally unity. $\Omega_0 = \Omega(r_0)$ is the rotation rate at the center of the channel. We introduce the Reynolds number $\reye = \Omega_0 r_0^2/\nu$, the magnetic Reynolds number $\reym = \Omega_0 r_0^2 / \eta$, and a plasma beta parameter $\beta = \Omega_0^2 r_0^2 \rho_0/B_0^2$. If we define the dimensional cylindrical coordinate $r = r_0(1 + \delta x)$, the thin gap equations are recovered in the limit $\delta \rightarrow 0$.

@@

\section{Weakly Nonlinear Perturbation Analysis}

@@

See Appendix \ref{app:basic_equations} for the definition of matrices and nonlinear vectors, and a thorough derivation. We emphasize that Equations \ref{eq:ordere} - \ref{eq:ordere3} have the same form as these equations in the thin gap case, although the matrices (which contain all radial derivatives) are significantly different in this wide gap formulation. This is because we do not have slow variation in the radial dimension. The slow variation in $Z$ and $T$ are parameterized as an amplitude function $\alpha(Z, T)$ which modulates the flow in these dimensions. This parameterization coupled with the boundary conditions lead us to an ansatz linear solution $\mathbf{V}_1 = \alpha(Z, T) \mathbb{V}_{11}(r) e^{i k_z z} + c.c.$, where the radial variation is contained in $\mathbb{V}_{11}$. We defer full expressions and derivations thereof to the Appendix, and focus here on features of the wide gap system which distinguish it from the thin gap limit.

@@

>>>>

\title{The Weakly Nonlinear Magnetorotational Instability in a Taylor Couette Flow}
\author{Clark, S.E.\altaffilmark{1}}
\author{Oishi, J.S. \altaffilmark{2,}\altaffilmark{3}}
%\author{Mac Low, M.-M.\altaffilmark{1,}\altaffilmark{2}}
\altaffiltext{1}{Department of Astronomy, Columbia University, New York, NY} 
\altaffiltext{2}{Department of Astrophysics, American Museum of Natural History, New York, NY}
\altaffiltext{3}{Department of Physics, SUNY Farmingdale}

@@

The magnetorotational instability (MRI) is thought to be a crucial driver of angular momentum transport and turbulence in astrophysical disks. (more MRI intro)

%The MRI is fundamental to our understanding of star formation, but its

Many investigations of the MRI use the ``narrow gap" approximation, in which the radial extent of the fluid channel is taken to be much smaller than the radius of curvature. That is, for a channel center $r_0$ bounded by inner and outer radii $r_1$ and $r_2$, respectively, the narrow gap approximation applies when $r_0 \gg (r_2 - r_1)$. The narrow gap approximation simplifies the MRI equations by excluding curvature terms, because the flow through a narrow gap can be taken to be approximately linear in $\phi$, i.e. Cartesian. In this work we undertake the first (to our knowledge) weakly nonlinear analysis of the MRI in the wide gap regime, where the channel width may be comparable to or larger than its distance from the center of rotation.

@@

@@

Thus through judicious choice of cylinder rotation rates, we can set $q(r_0) = 3/2$, for quasi-Keplerian flow. Note that in the narrow gap approximation we have the freedom to choose any rotation profile, as the narrow gap width imposes a linear shear (constant $q$). Our base velocity is $\mathbf{u_0} = r\Omega(r) \phihat$. We initialize an axial magnetic field $\mathbf{B_0} = B_0 \zhat$.

@@

where J is the Jacobian $J(x, y) \equiv \partial_z x \partial_r y - \partial_r x \partial_z y$. Note that in the above, $\nabla^2 x \equiv \partial_r^2 x + \partial_z^2 x + \frac{1}{r} \partial_r x$. Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are nondimensionalized, where lengths have been scaled by $r_0$, velocities by $r_0 \Omega_0$, densities by $\rho_0$, and magnetic fields by $B_0$; where $B_0$ appears in the above it is formally unity. $\Omega_0 = \Omega(r_0)$ is the rotation rate at the center of the channel. We introduce the Reynolds number $\reye = \Omega_0 r_0^2/\nu$, the magnetic Reynolds number $\reym = \Omega_0 r_0^2 / \eta$, and a plasma beta parameter $\beta = \Omega_0^2 r_0^2 \rho_0/B_0^2$. If we define the dimensional cylindrical coordinate $r = r_0(1 + \delta x)$, the narrow gap equations are recovered in the limit $\delta \rightarrow 0$.

@@

\section{Weakly Nonlinear Perturbation Analysis}\label{sec:wnl_analysis}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../python/widegap/figures/orders_1_2_widegap_fiducial_eigenfunctions.pdf}
\caption{Eigenfunctions}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../python/widegap/figures/nonlinear_2_3_widegap_fiducial_eigenfunctions.pdf}
\caption{Nonlinear terms}
\end{figure}

See Appendix \ref{app:basic_equations} for the definition of matrices and nonlinear vectors, and a thorough derivation. We emphasize that Equations \ref{eq:ordere} - \ref{eq:ordere3} have the same form as these equations in the narrow gap case, although the matrices (which contain all radial derivatives) are significantly different in this wide gap formulation. This is because we do not have slow variation in the radial dimension. The slow variation in $Z$ and $T$ are parameterized as an amplitude function $\alpha(Z, T)$ which modulates the flow in these dimensions. This parameterization coupled with the boundary conditions lead us to an ansatz linear solution $\mathbf{V}_1 = \alpha(Z, T) \mathbb{V}_{11}(r) e^{i k_z z} + c.c.$, where the radial variation is contained in $\mathbb{V}_{11}$. We defer full expressions and derivations thereof to the Appendix, and focus here on features of the wide gap system which distinguish it from the thin gap limit.

@@

Here we detail the perturbation analysis described in Section \ref{sec:wnl_analysis}. The linear system is described by Equation \ref{eq:unperturbed_matrix_eqns}, where

\beq
\mathcal{L} = \mathcal{L}_0 + \mathcal{L}_1 \partial_z + \mathcal{L}_2 \partial_z^2 + \mathcal{L}_3 \partial_z^3 + \mathcal{L}_4 \partial_z^4,
\eeq

\beq
\widetilde{\mathcal{G}} = - \mathcal{G} \partial_z - \mathcal{L}_3 \partial_z^3,
\eeq

\beq
\widetilde{\mathcal{H}} = \mathcal{H} \partial_z,
\eeq

and the constituent matrices are defined as

\beq
\mathcal{L}_0 = \left[\begin{matrix}
-\frac{1}{\reye} (-\frac{3}{r^4} \partial_r + \frac{3}{r^3}\partial_r^2 - \frac{2}{r^2}\partial_r^3 + \frac{1}{r}\partial_r^4) & 0 & 0 & 0 \\
0 & -\frac{1}{\reye} (\partial_r^2 + \frac{1}{r}\partial_r - \frac{1}{r^2}) & 0 & 0 \\
0 & 0 & -\frac{1}{\reym} (\partial_r^2 - \frac{1}{r} \partial_r) & 0 \\
0 & 0 & 0 & -\frac{1}{\reym} (\partial_r^2 + \frac{1}{r}\partial_r - \frac{1}{r^2}) \\
\end{matrix}\right]
\eeq

\beq
\mathcal{L}_1 = \left[\begin{matrix}
0 & -\frac{2}{r} u_0 & \frac{2}{\beta} (\frac{1}{r^2} \partial_r - \frac{1}{r}\partial_r^2) & 0 \\
\frac{1}{r^2} u_0 + \frac{1}{r}\partial_r u_0 & 0 & 0 & -\frac{2}{\beta} \\
-1 & 0 & 0 & 0 \\
0 & -1 & \frac{1}{r^2} u_0 - \frac{1}{r} \partial_r u_0 & 0 \\
\end{matrix}\right]
\eeq

\beq
\mathcal{L}_2 = \left[\begin{matrix}
-\frac{1}{\reye} (-\frac{2}{r^2}\partial_r + \frac{2}{r}\partial_r^2) & 0 & 0 & 0 \\
0 & -\frac{1}{\reye} & 0 & 0 \\
0 & 0 & -\frac{1}{\reym} & 0 \\
0 & 0 & 0 & -\frac{1}{\reym} \\
\end{matrix}\right]
\eeq

\beq
\mathcal{L}_3 = \left[\begin{matrix}
0 & 0 & -\frac{2}{\beta}\frac{1}{r} & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
\end{matrix}\right]
\eeq

\beq
\mathcal{L}_4 = \left[\begin{matrix}
-\frac{1}{\reye}\frac{1}{r} & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
\end{matrix}\right]
\eeq

\beq
\mathcal{G} = \left[\begin{matrix}
0 & 0 & \frac{2}{\beta}(\frac{1}{r^2}\partial_r - \frac{1}{r}\partial_r^2) & 0 \\
0 & 0 & 0 & -\frac{2}{\beta} \\
-1 & 0 & 0 & 0 \\
0 & -1 & 0 & 0 \\
\end{matrix}\right]
\eeq

\beq
\mathcal{H} = \left[\begin{matrix}
0 & 0 & 0 & \frac{2}{\beta} \frac{2}{r^2} \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
-\frac{2}{r^3} & 0 & 0 & 0 \\
\end{matrix}\right]
\eeq
